---
title: "A Tutorial to SpatialROI"
author: "Mengyao Lu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpatialROI User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting Started with SpatialROI

SpatialROI is an interactive web application for region-specific analysis and visualization of spatial transcriptomics data. This vignette provides a step-by-step walkthrough of SpatialROI's functionality using the built-in example dataset.

## Launching the Application

To launch SpatialROI, run the following command in R:

```{r eval=FALSE}
library(SpatialROI)
run_SpatialROI("demo")
```

The application will open in your default web browser. You can simply click "Start Analysis" in the main intro page.

---

## Quick ROI Selection with `draw_ROI()`

**For users who need immediate spot selection without the full analysis interface:**

The `draw_ROI()` function provides a lightweight interactive tool for quickly selecting regions of interest and obtaining spot IDs for downstream analysis.

### Usage

```{r eval=FALSE}
library(SpatialROI)

# Load your data
my_data <- readRDS("path/to/your_seurat.rds")

# Launch interactive ROI selector
selected_spots <- draw_ROI(my_data, sample_name = "MyExperiment")

# Use the returned spot IDs
print(length(selected_spots))  # Number of spots selected
subset_data <- subset(my_data, cells = selected_spots)
```

### Key Features

- **🖊️ Freehand drawing** - Click the ✏️ pencil button to draw custom irregular regions
- **📏 Interactive controls** - Adjust spot size and H&E tissue opacity in real-time
- **💾 Immediate export** - Returns spot IDs directly for use in your analysis pipeline

### When to Use Each Tool

| Tool | Best For | Output |
|------|----------|--------|
| `draw_ROI()` | Quick spot selection, multiple ROI extraction, preprocessing | Spot ID vector |
| `run_SpatialROI("demo")` or `run_spatial_selector(object)` | Full analysis: DEGs, clustering, comparisons, gene sets | Interactive analysis app |

**Note:** `draw_ROI()` returns spot IDs for lightweight scripted use. The full
SpatialROI app treats every saved ROI as a named region and optionally combines
multiple ROIs into a named group. Both ROI-versus-ROI and group-versus-group
comparisons are available from the same region selectors.

---

## Data Loading

SpatialROI provides two convenient options for loading data: using a built-in demo dataset or uploading your own Seurat object.

### Option 1: Using Demo Dataset (Recommended for First-Time Users)

You can launch the app with the preloaded example dataset by running:

```{r eval=FALSE}
library(SpatialROI)
run_SpatialROI("demo")
```

#### Example Data

The demo dataset is one human colorectal cancer Visium tissue section from
Valdeolivas et al. (2024), containing 17,529 genes and 1,253 tissue spots, an H&E
image, and a preprocessed Seurat object. Users can load this example for testing
or demonstration purposes.

### Option 2: Uploading Your Own Data

If you have your own spatial transcriptomics dataset (e.g., a .rds or .rdata file containing a Seurat object), you can launch the app with it:

```{r eval=FALSE}
library(SpatialROI)
my_data <- readRDS("path/to/your_seurat.rds")
run_spatial_selector(my_data, sample_name = "MyExperiment")
```

Alternatively, after launching the app with the demo data, you can:

1. Go to the **"Visualization"** tab on the left in the app interface.
2. On the right panel, click **"Upload Data"** to select and upload your .rds file.
3. After the "Upload complete" message appears, click **"Load Uploaded Data"**.
4. This will replace the current dataset and begin analysis on your own data.

**Note:** Uploading new data will reset any existing ROIs or analyses from the current session.

---

## Visualization

Once your dataset is loaded, you'll be directed to the interactive visualization page, where the H&E image is displayed with spatial spots overlaid. This section introduces the layout and basic tools for customizing your view before analysis.

### Interface Overview

At the top left of the image panel, there are **"+"** and **"−"** buttons for zooming in and out, and a **pen icon** that activates ROI selection mode. You'll see your tissue H&E image displayed in the main center panel, with circular spot overlays indicating transcriptomic measurements.

### Zoom and Pan

- **Zoom In/Out:** Use your mouse scroll wheel over the image, or use the zoom buttons (**+** and **−**) in the top-left corner of the image panel.
- **Pan:** Click and drag anywhere on the image to move across different tissue regions.

### Adjusting Spot Size

1. Locate the **"Spot Size"** slider in the top right sidebar.
2. Drag left to decrease spot size (useful when zoomed out or for dense arrays).
3. Drag right to increase spot size (helpful when zoomed in or for sparse arrays).

The default setting works well for most Visium arrays but can be adjusted for visual clarity.

### Adjusting H&E Opacity

1. Locate the **"H&E Opacity"** slider in the top right sidebar.
2. Drag the slider left to make the H&E image more transparent (0% = fully transparent).
3. Drag the slider right to make it more opaque (100% = fully opaque).

This allows better visibility of the overlaid spatial spots. **Recommended setting:** 60–80% for optimal balance between tissue detail and spot visibility.

---

## Interactive ROI Selection

This is one of SpatialROI's key features, allowing you to define regions of interest (ROIs) directly on tissue images. This is useful for grouping spatial regions for downstream comparison.

### Drawing ROIs

1. **Activate Drawing Mode:**  
   Click the **pen icon** at the top-left of the image panel to enter ROI selection mode. A toolbar will appear with freeform drawing tools.

2. **Draw and Save a Named ROI:**
   Draw an irregular region. To enhance visibility, reduce the **Spot Size** and increase the **H&E Opacity**. The selected spots will turn yellow.  
   Enter a biologically meaningful name (for example, `Tumor edge 1`) and click
   **Save region**. Repeat this step for each distinct component you want to keep.

3. **Optionally combine ROIs into groups:**
   Select two or more saved ROIs, enter a group name (for example,
   `Tumor-adjacent`), and click **Group**. A group is the union of its component
   ROIs; the original ROIs remain available for ROI-level analysis.

4. **Visualize ROIs or groups on the map:**
   Choose an ROI, a group, or all ROIs under **Show on map**, then enable
   **Show ROIs on Map**. Enable **Show ROI contours** for high-contrast boundaries.

5. **Reset or Redraw:**
   Click **"Clear Selection"** to clear all highlighted spots and start a new ROI selection.

6. **Export a region:**
   Select a saved ROI or group under **Show on map**, then use **Save region
   (.rds)** to export the corresponding Seurat subset.

### Two comparison designs

**Example A — ROI versus ROI.** Draw and save two single regions, such as
`TLS-like region` and `Tumor region`. In the DEG or Compare panel, set these as
Side A and Side B. Use this design when each boundary itself represents the
biological unit of interest.

**Example B — group versus group.** Draw several separate tumor-edge regions and
save them as `Tumor edge 1`, `Tumor edge 2`, and so on; combine them into a named
`Tumor-adjacent` group. Repeat for several normal-edge regions and combine them
as `Normal-adjacent`. Compare the two named groups. Grouping unions the selected
spots for a within-section exploratory comparison; it does not create additional
patient-level replicates.

---

## Visualization of Features and Gene Set Scoring

SpatialROI allows users to explore and visualize spatial gene activity through both **single-feature mapping** and **gene set scoring**.

### Visualizing Features

After loading your data, navigate to the **"Feature Selection"** panel to explore spatial features.

#### Feature Selection

- Under **Feature Type**, choose from:
  - **None** — disables feature display
  - **Gene Expression** — visualize the spatial distribution of a specific gene
  - **Metadata** — show spatial patterns of any sample or cell-level metadata
- When **Gene Expression** is selected, a dropdown of available genes will appear. Selecting one gene or typing in one gene name will automatically render its expression overlay on the H&E image.

#### Color Scheme

- You can customize how features are visualized using the **Color Scheme** section.
- Available palettes include:
  - *Grey to Red*
  - *Blue–White–Red*
  - *Rainbow*
- The selected color gradient will immediately apply to the expression overlay, allowing users to interpret expression intensity spatially.

### Gene Set Scoring

SpatialROI also supports **gene set based analysis** to visualize composite activity scores from curated or custom gene lists.

#### Species Selection

- Choose between **Human** or **Mouse** from the **Select Species** dropdown.
- Gene symbols will be automatically updated to match the selected species.

#### Selecting Gene Sets

- Under **Select Signature**, you can choose from:
  - **Predefined Signatures** curated from the **CellMarker 2.0** database (Hu *et al.*, *NAR* 2023), including immune, stromal, and cancer-related markers such as *immune: B cells* or *cancer: epithelial*.
  - After selecting a predefined signature, click **Load Selected Signature** to populate the gene list automatically. You can freely edit or add genes afterward.
  - **Custom Gene Sets** — Manually input/delete genes in the **Gene Input** box (one gene per line).

#### Scoring Methods

- In the **Parameters** section, select your preferred scoring method:
  - **Mean** — computes the average expression of all genes in the set
  - **AddModuleScore** — implements Seurat's control-feature–adjusted score
  - **Rank-based enrichment** — computes a within-spot mean-rank difference,
    scaled approximately to [-1, 1]. This is implemented directly by SpatialROI
    and is **not** the GSVA package statistic.
- Select a color scheme (*Grey to Red*, *Blue–White–Red*, or *Rainbow*).
- Click **Calculate Score** to compute scores. Once finished, results will appear as an overlay on the main H&E slide, highlighting regions with enriched expression.

#### Saving and Reusing Gene Sets

- After scoring, click **Save to Gene Set** to store the calculated results for downstream analyses or re-visualization.

SpatialROI thus integrates flexible **feature-level** visualization with **multi-gene set analysis**, enabling users to map spatial expression patterns and functional gene activities directly onto tissue histology for biologically meaningful interpretation.

---

## ROI-based Clustering

Use ROI-based clustering to identify transcriptionally distinct regions within your tissue.

### Running Clustering Analysis

1. Go to the **"Clusters"** tab.
2. Under **"Cluster which spots"**, choose the scope of analysis:
   - **All Spots** — cluster all spots in the tissue
   - **Any saved ROI** — cluster only that morphology-defined region
   - **Any saved group** — cluster the union of its component ROIs

3. Set clustering parameters:
   - **Resolution** controls the number of clusters (range 0.1–2.0, default 0.8).
   - **Principal Components (PCs)** specify how many components are used for clustering (default 30).

4. Click **"Run Clustering"** to start. The system will process your data and assign each spot to a cluster.

### Visualizing Clusters

When clustering is complete, results will appear on the right panel, showing the selected scope, number of spots, resolution, number of clusters, and the spot count in each cluster. A **UMAP** plot will also be displayed for an overview of cluster separation.

Click **"Show on Map"** to overlay the cluster results on the H&E image. Each cluster will appear in a different color, and unselected spots will turn grey.

The clustering results are automatically saved in SpatialROI for further analysis, such as differential expression or ROI comparison.

---

## Differential Expression Analysis

Identify genes that are differentially expressed between any two saved ROIs or
groups, or between one saved region and the rest of its tissue section.

### Setting Up the Comparison

1. Go to the **"DEGs"** tab.

2. Select **Side A** as any ROI or group. Select **Side B** as another ROI/group
   or **Rest of tissue**. This supports ROI-versus-ROI, group-versus-group,
   mixed ROI-versus-group, and region-versus-rest comparisons.

3. Click **"Find DEGs"** to start.  
   SpatialROI automatically runs Seurat's `FindMarkers` function and returns the **top differentially expressed genes** ranked by *p-value*. Each result includes the gene name, average log2 fold change (log2FC), and adjusted p-value (*p_adj*).

   When comparing clusters (e.g., one vs rest), the output also includes:
   - **pct.1**: percentage of spots expressing the gene in the target cluster
   - **pct.2**: percentage of spots expressing the gene in all other clusters
   
   These help indicate how specific the gene is to each cluster.

4. To download all DEGs, click **"Download DEG Results"** to save the full results table as a CSV file.

### Viewing Results

- The results table shows:
  - Gene name
  - Average log2 fold change
  - Adjusted p-value
  - pct.1 and pct.2 (proportion of spots expressing the gene in each group)
- You can sort or search the table, and export results for further analysis.

---

## Feature or Group Comparison Across Regions

Compare the expression of a single feature, gene set, or metadata variable between regions.

### Single Feature Group vs Group Comparison

1. Go to the **"Compare"** tab, then select the **Feature** panel on the right.

2. Under **Feature Type**, choose one of the following:
   - **Gene** — compare expression of a single gene
   - **Metadata** — compare continuous or categorical variables
   - **Gene Set** — compare pre-defined or custom gene set scores

3. Select any saved ROI or group for **Side A**, and any ROI/group or the rest
   of tissue for **Side B**.

4. Choose the **Statistical Test**:
   - **t-test** — compares group means (use for normally distributed data)
   - **Wilcoxon/Mann–Whitney** — non-parametric alternative (use when data are not normally distributed)

5. Click **"Generate Plot"**.  
   A violin plot will display the distribution of the selected feature across regions. The number of spots and p-values from the statistical test will appear on the plot.

### Feature vs Feature Comparison Within the Same Region

1. Select two features to compare (Feature 1 and Feature 2).
2. Choose the spot population under **Use Spots**: all spots, any saved ROI, or
   any saved group.
3. Pick the desired statistical test (t-test or Wilcoxon/Mann–Whitney) and click **"Compare"**.  
   A violin plot will show the distributions of the two features in the selected region, along with p-values and sample counts.

This function helps identify spatial differences in gene expression, module scores, or metadata between regions or within clusters, offering biological insight into spatial heterogeneity across the tissue.

---

## Tips for Optimal Results

### ROI Selection Best Practices

- **Start broad, then refine**: Create larger ROIs first to capture main regions, then zoom in to define sub-regions
- **Use appropriate H&E opacity**:
  - 70-80% when you need to see tissue morphology clearly
  - 40-60% when you need to see spot locations clearly
  - Toggle opacity as you draw to ensure accurate selection
- **Adjust spot size based on zoom level**:
  - Smaller spots (2-4) when zoomed out for overview
  - Larger spots (6-10) when zoomed in for precise selection

### Visualization Tips

- **Color scheme selection**:
  - Among the available palettes (Grey to Red, Blue–White–Red, Rainbow), we recommend using Blue–White–Red for clearer and more interpretable contrast
  - If you visualize a single gene and later a gene set, the new visualization will replace the previous one by default
  - For better interpretability, consider using different color schemes for different visualization types (e.g., Blue–White–Red for gene expression, Rainbow for gene set scores)
- **Interactive exploration**:
  - Use zoom and pan freely to inspect spatial regions of interest in greater detail
  - You may also consider loading the same tissue image into QuPath or other specialized digital pathology tools to compare ROIs and enhance visualization quality or annotation capabilities

### Analysis Guidelines

- **Preprocessing**:
  - Filter low-quality samples and rare taxa before downstream analysis
  - Consider your biological question and assumptions when choosing your ROIs
- **Gene set scoring**:
  - Use Seurat's AddModuleScore for relative comparisons
  - Use GSVA when absolute scores are important
  - Validate interesting patterns with individual gene expression
- **Differential expression**:
  - Use appropriate fold change thresholds based on your data type
  - Check that both groups have sufficient cell numbers (>30 spots recommended)
  - Validate top hits with spatial visualization and later comparison
- **Statistical testing**:
  - Use non-parametric tests for small sample sizes
  - Consider biological significance, not just statistical significance

---

## Troubleshooting

### Common Issues and Solutions

**Data loading fails:**

- Ensure your files are in the correct format (.rds or .RData)
- Verify that spatial coordinates are numeric
- Try uploading files individually to identify the problematic file

**Seurat object version mismatch error:**

If you encounter an error like `Error: invalid class "VisiumV1" object: slots in class definition but not in object: "misc"`, this indicates a version mismatch between the Seurat version used to create your object and your current Seurat installation. To fix this:
```r
# Update your Seurat object
old_obj <- readRDS("path/to/your_seurat.rds")
new_obj <- UpdateSeuratObject(old_obj)
saveRDS(new_obj, "path/to/your_seurat.rds")
```

Then reload the updated file in SpatialROI.

**ROIs not appearing on map:**

- Increase spot size using the slider
- Adjust H&E opacity to ensure spots are visible
- Check that the coordinate system matches your image
- Try refreshing the page and reloading data

**Analysis runs slowly:**

- Reduce the number of features analyzed
- Filter low-quality data more stringently
- Close other programs to free up memory

**Visualizations appear distorted:**

- Reset zoom to home position
- Adjust spot size and opacity
- Check browser zoom level
- Try a different browser

**Can't find a specific gene:**

- Check gene name spelling (mouse/human)
- Verify the gene is present in your dataset
- Some genes may be filtered during preprocessing
- Confirm the species (human/mouse)

---

## Getting Help

If you encounter issues not covered in this vignette:

Visit the SpatialROI GitHub repository for:

- Detailed documentation
- Example datasets
- Issue reporting

---

## Citation

If you use SpatialROI in your research, please cite:

[Citation here]

---

## Conclusion

This vignette has walked you through all major features of SpatialROI, from data loading through analysis and export. The interactive nature of SpatialROI makes it a powerful tool for exploratory analysis of spatial transcriptomics data, allowing you to identify biologically relevant regions and understand spatial gene expression patterns.

Happy analyzing!
