---
title: "A Tutorial to SpatialROI"
author: "Mengyao Lu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpatialROI User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

SpatialROI is a Shiny application for region-based analysis of 10x Visium data.
You draw regions of interest (ROIs) on the H&E image, then analyse and compare
them without writing code. This guide follows the app from left to right in the
sidebar. The hosted version is at <https://shiny.crc.pitt.edu/spatial_api/>.

## Installation and launch

```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("Seurat", "GSVA"))
devtools::install_github("myaol/SpatialROI", dependencies = TRUE)

library(SpatialROI)
run_SpatialROI("demo")                      # start with the bundled CRC section

my_data <- readRDS("path/to/your_seurat.rds")
run_spatial_selector(my_data, sample_name = "MyExperiment")   # start with your own object
```

Cell-type deconvolution needs `spacexr`, and differential expression runs much
faster with `presto`. Both are installed from GitHub:

```{r eval=FALSE}
remotes::install_github("dmcable/spacexr@698d5b09e6f70d75d7a5284cf90f8ba4adcb7205")
remotes::install_github("immunogenomics/presto")
```

A local installation accepts uploads up to 500 MB. To raise this, set
`options(shiny.maxRequestSize = ...)` before launching. The hosted server
accepts up to 150 MB.

## 1. Load data (Visualization)

Click **Start Analysis** on the home page, then open **Visualization** in the
sidebar. Under **Data Source**, choose one of two input types:

- **10x Visium Seurat object (.rds)**: a processed object with a spatial image,
  spot coordinates and log-normalised expression. Upload it and click
  **Load Seurat Object**. Two bundled datasets are available here:
  **Default Data (CRC)**, a colorectal cancer section with 1,253 spots, and
  **Case Study 1 (CRLM)**, a colorectal cancer liver metastasis with 3,721 spots.
- **10x Visium Space Ranger output (.zip)**: the Space Ranger output folder,
  unzipped files only (no `.gz`), containing the `.h5` matrix and the `spatial/`
  folder, compressed as one `.zip`. Spots are filtered with the QC settings
  (minimum genes, minimum counts, maximum % mitochondrial) and then normalised.
  **Case Study 2 (OSCC)** is bundled in this format.

Loading new data replaces the current analysis, including all ROIs.

## 2. Explore the map

Zoom with the scroll wheel or the **+/−** buttons, and drag to pan. **Spot Size**
and **H&E Opacity** control how the spots sit over the tissue.

Under **Feature Selection**, choose **Gene Expression** or **Metadata** to colour
the spots, and pick a palette under **Color Scheme** (Grey to Red, Blue-White-Red
or Rainbow). **Clear Map Overlay** removes any feature, score or cluster colouring.
**Save view (PDF)** exports the current map.

## 3. Draw regions and groups

1. Click the pencil icon on the map and draw a region freehand. Smaller spots and
   higher H&E opacity make the tissue easier to see while drawing.
2. Name the region and click **Save region**. Repeat for each region you need.
3. To analyse several ROIs together, select them, enter a group name and click
   **Group**. A group is the union of its ROIs, and the ROIs stay available on
   their own.
4. Use **Show on map** with **Show ROIs on Map** to display a region.
   **Transparent ROI Display** shows the tissue underneath, and
   **Show ROI contours** draws a high-contrast outline.

Every saved ROI or group can be used in each analysis panel. Comparisons can be
ROI vs ROI, group vs group, ROI vs group, or a region vs the rest of the tissue.
Grouping pools spots within one section. It does not create biological replicates.

**Save and reload regions.** **Save ROI index (.csv)** writes the spot IDs of a
region. **Load ROI index (.csv)** reads one back in a later session, and redraws
its outline on the map. The file only needs a `spot_id` column, so regions
selected in other tools can be imported the same way. **Save region (.rds)**
exports the selected region as a Seurat object. The spot indices of the TLS
regions used in the manuscript are in `inst/extdata/example_rois/`.

## 4. Ligand–receptor colocalisation

In **Visualization**, the **L-R Colocalization Score** section ranks
ligand–receptor pairs within all spots, one ROI, or one group. Select the species
and the databases to include (CellChat, KEGG, Guide to Pharmacology and
Ramilowski, all selected by default), then click **Run LR Scoring**.

For each gene, expression is smoothed over each spot and its 12 nearest
neighbours with a Gaussian kernel. The **Gaussian bandwidth multiplier** sets how
wide that kernel is. A pair's score at a spot is the geometric mean of its
smoothed ligand and receptor expression. The table reports the mean of that score
over the spots in the region. The databases only decide which pairs are scored.
The scores come entirely from your data. Highly ranked pairs are candidates for
local signalling, not evidence of it.

## 5. Cell-type deconvolution

The **Cell Type Deconvolution** section estimates the cell-type composition of
each spot with RCTD (`spacexr`, full mode).

1. Choose a reference. Two are built in: **CRC** (GSE132465), which matches
   Default Data, and **CRLM** (GSE225857), which matches Case Study 1. Loading a
   bundled dataset pre-selects its matching reference. The CRLM reference has no
   hepatocytes, so it suits tumour and tumour-adjacent regions rather than normal
   liver. Alternatively, upload your own single-cell Seurat object (.rds) with
   raw counts and cell-type labels, either as identities or in a metadata column.
2. Pick one or more regions under **Regions to deconvolve**, or **All spots**.
3. Click **Estimate Cell Composition**.

RCTD runs in a background process, so the rest of the app stays usable. Large
regions take several minutes. Each region is fitted separately. To compare
proportions strictly across regions, deconvolve **All spots** once and compare
within that run. Proportions are relative to the cell types in the reference.
A cell type missing from the reference is assigned to the ones that are there.
Curated references for other cancer types are listed on the GitHub page.

## 6. Gene set scoring (Gene Sets)

Select the species first. Then load genes from one of three sources:

- **Marker Signatures (CellMarker 2.0)**: immune, stromal, tissue and cancer
  signatures. Click **Load Selected Signature**.
- **Pathway Signatures (MSigDB Hallmark)**: click **Load Pathway Genes**.
- Your own list, typed into **Gene Input** (one gene per line or comma-separated).

The list can be edited after loading. Choose a **Method**:

- **Mean expression**: the average expression of the genes.
- **AddModuleScore (Seurat)**: the average minus that of matched control genes.
- **GSVA (rank-based)**: the `GSVA` package score.

Click **Calculate Score** to display the score on the map. **Save to gene set**
keeps it under the given **Name**, so it can be used in the Compare panel.

## 7. Clustering (Clusters)

Choose **Cluster which spots**: all spots, an ROI, or a group. Set the
**Resolution** (default 0.8) and the number of **PCs** (default 30), then click
**Run Clustering**. The panel reports the number of clusters and spots per
cluster, and shows a UMAP. **Show on Map** colours the spots by cluster. The UMAP
and the cluster assignments can be downloaded.

## 8. Differential expression (DEGs)

Set **Side A** to an ROI or group, and **Side B** to another ROI or group or to
**Rest of tissue**. Non-overlapping regions are recommended. If they overlap,
shared spots are left out of both sides.

Choose the **DE test** (Wilcoxon, t-test or logistic regression), the
**|log2FC|** cut-off, the **Min. spot fraction** and the **Adjusted p (BH)**
threshold, then click **Find DEGs**. Seurat's `FindMarkers` tests every gene that
passes the spot-fraction filter, and p-values are adjusted with Benjamini–Hochberg
across all tested genes. The fold-change cut-off is applied afterwards and does
not change the adjusted p-values.

The volcano plot shows all tested genes. The table, and the downloaded CSV,
contain only the genes that pass both thresholds. Moran's I, computed over the
whole section, shows which of those genes are spatially structured. A table of
one region vs the rest of the tissue can be uploaded to the Multi-Sample panel.

## 9. Comparisons (Compare)

**Group vs Group** compares one feature between two regions. The feature can be a
gene, a metadata column, a saved gene set, a Hallmark pathway or a cell-type
signature. Choose **Side A** and **Side B** and a test (Wilcoxon or t-test), then
click **Generate Plot**. The result is a violin plot with the p-value and the
number of spots.

**Feature vs Feature** compares two features within the same spots: all spots,
an ROI, or a group. Choose the test, then click **Generate Plot**.

## 10. Comparing ROIs across samples (Multi-Sample)

This panel compares DEG tables rather than expression data, so ROIs from
different sections or studies can be compared without integrating them. Each
table should be one region vs the rest of its tissue. Tables from SpatialROI can
be uploaded directly. Tables from other tools need a gene column (such as `gene`
or `symbol`) and a log-fold-change column (such as `avg_log2FC`,
`log2FoldChange` or `logFC`). **Load example tables** loads three TLS tables from
the bundled datasets.

1. **Upload DEG tables**: at least two. The summary shows each table's size and
   the number of up- and down-regulated genes.
2. **Compare ROI similarity**: for each pair of tables, over the genes reported in
   both, the number of shared genes, the percentage that change in the same
   direction, and the Spearman correlation of their log2 fold changes.
3. **Compare differential expression patterns across ROIs**: a heatmap of genes
   reported in every table. Half the rows are the strongest shared up-regulated
   genes and half the strongest shared down-regulated genes. The table below it
   lists every shared gene.
4. **Compare pathway enrichment across ROIs**: Hallmark over-representation, run
   separately for each table, with all Hallmark genes as the background.
   Pathways enriched in the most ROIs come first. The species follows the setting
   in the Gene Sets panel.

These summaries are descriptive. Several ROIs from one patient are not
independent replicates. Adding more tables usually reduces the number of genes
shared by all of them.

## Troubleshooting

**An older Seurat object fails to load** with an error such as
`invalid class "VisiumV1" object`. Update it and save it again:

```{r eval=FALSE}
obj <- UpdateSeuratObject(readRDS("path/to/your_seurat.rds"))
saveRDS(obj, "path/to/your_seurat.rds")
```

**A gene cannot be found.** Check the species setting and the gene symbol.
The gene may also have been removed during preprocessing.

**Deconvolution reports few matched genes.** The reference and the spatial data
must use the same species and gene symbols.

**The session was lost.** Uploaded data is kept only while the browser tab is
open. Save ROI indices and results before closing the tab.

## Scripted ROI selection

`draw_ROI()` opens a drawing map in the browser and returns the spot IDs of the region
you draw. This is useful for scripted workflows:

```{r eval=FALSE}
spots <- draw_ROI(my_data, sample_name = "MyExperiment")
roi   <- subset(my_data, cells = spots)
```

## Citation

SpatialROI is described in a manuscript in preparation. Citation details will be
added here on publication. Issues and questions can be posted on
<https://github.com/myaol/SpatialROI>.
